Bayesian Restricted Likelihood Methods

### 2018-07-09 submission_1/BlendedParadigm0832.tex
gathering all revision in order to work on paper again

% commands for for editing
\usepackage{color}
\usepackage{ulem} % \sout
\newcommand{\red}[1]{{\color{red}#1}}
\newcommand{\blue}[1]{{\color{blue}#1}}
\newcommand{\brown}[1]{{\color{brown}#1}}
\newcommand{\green}[1]{{\color{green}#1}}
\newcommand{\magenta}[1]{{\color{magenta}#1}}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\graphicspath{{figures/}}

@@

%\newpage
\title{Bayesian Restricted Likelihood Methods: Conditioning on Insufficient Statistics in Bayesian Regression}
\author{John R. Lewis, Steven N. MacEachern and  Yoonkyung Lee \\
{\small \it Department of Statistics, The Ohio State University, Columbus, Ohio 43210}\\
{\small lewis.865@osu.edu, snm@stat.osu.edu and yklee@stat.osu.edu}
\thanks{This research has been supported by Nationwide Insurance Company and by the NSF under grant numbers DMS-1007682 and DMS-1209194.  The views in this paper are not necessarily those of Nationwide Insurance or the NSF.}} 
\begin{document}
\date{}
\maketitle

\begin{abstract}
Bayesian methods have proven themselves to be successful across a wide
range of scientific problems and have many well-documented advantages
over competing methods. However, these methods run into difficulties
for two major and prevalent classes of problems: handling data sets
with outliers and dealing with model misspecification. We outline the
drawbacks of previous solutions to both of these problems (e.g., use of
heavy-tailed likelihoods) and propose a new method as an
alternative.  When working with the new method, we summarize
the data through a set of insufficient statistics,
targeting inferential quantities of interest, and update the prior
distribution with the summary statistics rather than the complete
data.  By careful choice of conditioning statistics, we
retain the main benefits of Bayesian methods while reducing the
sensitivity of the analysis to features of the data not captured by
the conditioning statistics. For reducing sensitivity to outliers,
classical robust estimators (e.g., M-estimators) are natural choices
for conditioning statistics. With these choicees, the method can be thought of as a
blend of classical robust estimation and Bayesian methods.

A major contribution of this work is the development of a data 
augmented Markov chain Monte Carlo (MCMC) algorithm
for the linear model and a wide range of choices for
summary statistics. We demonstrate the method on an
insurance agency data set containing many outliers and subject to model
misspecification. Success is manifested in better predictive
performance for data points of interest as compared to competing
methods.

@@

\section{Introduction}
Bayesian methods have provided successful solutions to a wide range of scientific problems, with their value
having been demonstrated both empirically and theoretically.  
In simple settings, the success of the methods is often attributed to formal optimality properties, sometimes derived
through the laws of subjective probability 
and sometimes through admissibility and the complete class
theorems of decision theory.  In complex settings, the hierarchical model allows one to create 
and fit sophisticated models that may, for example, pool information across similar problems.

The development of Bayesian inference relies on a complete Bayesian model consisting of 
three elements:  the prior distribution, the loss function, and the likelihood or sampling density.  While formal
optimality of Bayesian methods is unquestioned if one accepts the validity of all three of these elements, 
 a healthy skepticism encourages us to question each of them.  Concern about the prior distribution has
been addressed through the development of techniques for subjective elicitation \citep{garthwaite2005, ohagan2006} \blue{(added the O'Hagan text here)} and
the rise of objective Bayesian methods \citep{berger2006}.  
Concern about the loss function is reflected in, for example, the
extensive literature on Bayesian hypothesis tests \citep{kass1995}.  
The sampling density has been given less attention from a specifically
Bayesian view, although the work on predictive diagnostics \citep{box1980} departs from classical
traditions.

%The focus of this work is the intersection of Bayesian methodology and data analysis. In particular, we develop techniques to handle imperfections in the sampling density
The focus of this work is the development of techniques to handle imperfections in the sampling density.  These imperfections often show
themselves through the presence of outliers--cases %--here taken to be 
not reflecting the phenomenon under
study. %--in the data set. 
%A traditional solution to Bayesian outlier-handling involves replacing the original sampling density with a mixture or heavy-tailed density.  A mixture model  includes one component for the ``good'' data and a second component for the ``bad'' data in an attempt to compartmentalize the outliers.  The heavy-tailed approach simply attempts to discount the outliers directly. Another traditional approach fits a flexible (typically nonparametric) model to the data, producing a Bayesian version of a density estimate for both good and bad data.  In recent development, inference is made through the use of robust inference functions \citep[][in press]{lee2013}.  
%\red{Q: Is the above sentence still about modifying the sampling density to
%  handle outliers?}
There are three main solutions to Bayesian outlier-handling.  The first is to replace the basic sampling
density with a mixture model which includes one component for the ``good'' data and a second 
component for the ``bad'' data.  With this approach, the prior distribution is updated with the likelihood
from the mixture model to obtain the complete-data posterior distribution.  The good component of
the sampling density is used for prediction of future good data.  The second approach replaces the
basic sampling density with a thick-tailed density in an attempt to discount outliers, yielding techniques that 
often provide solid estimates of the center of the distribution but 
do not easily translate to predictive
densities for further good data.  The third approach fits a flexible (typically nonparametric) model to 
the data, producing a Bayesian version of a density estimate for both good and bad data.  In recent 
development, inference 
is made through the use of robust inference functions \citep[][in press]{lee2013}.

The traditional strategies for handling outliers all have their drawbacks.  While we view the sampling 
density for the good data as stable, the outlier-generating processes 
may be transitory in nature, constantly shifting as the source of bad data changes.  This prevents
us from appealing to large-sample arguments to claim that, with enough data, we can nail down a
model for both good and bad data combined.  Instead of attempting to model both good and bad
data, we propose a novel strategy for handling outliers. %:  the use of restricted likelihood.  
In a nutshell, we begin with a complete model   
as if all of the data are good.
But rather than driving the move from prior distribution to posterior
distribution by the entire likelihood, we use only the likelihood
driven by a
few summary statistics which typically target inferential quantities
of interest.  We call this reduced likelihood a restricted % an incomplete
likelihood. 
%\red{[better to call a restricted likelihood? if we change, make sure to change consistently throughout the paper]} 
%\green{[I agree with Yoon.  Better to call this a restricted likelihood.]}
The 
update is a formal update from prior distribution to posterior distribution, based on the sampling density
of the summary statistics. 
 %In our approach, the reader will identify a stream of reasoning which is manifested in classical M-estimation, generalized estimating equations, approximate Bayesian computation, and elsewhere.  
The novelty of the work is twofold.  We make use of
classical robust estimators as  summary statistics in a formal
Bayesian framework, using the sampling density of the estimators as a replacement for the 
sampling density of the data.  Second, we advance the argument that conditioning on an insufficient
summary of the data is sound practice, rather than merely being done for computational and 
modelling convenience.

% \red{[Steve, just a reminder that a paragraph about model misspecification was here orginally but commented out by John following a referee's suggestion. I also agree with the referee in that model misspecification could be a much broader issue than having a fraction of bad data while the model for good data is correctly specified.]}
%In addition to outlier-prone data, the 
\green{ The sampling density can err due to other forms of model misspecification.  
The traditional view is that, if the model is inadequate, one should build a better model.  In our
empirical work, as data sets have become larger and more complex, we have bumped into 
settings where we cannot realistically build the perfect model.  We ask the question ``by attempting
to improve our model through elaboration, will the overall performance of the model suffer?''  If yes,
we avoid the elaboration, retaining a model with some level of misspecification.  Acknowledging that
the model is misspecified implies acknowledging that the sampling density is incorrect, as
we do when outliers are present.  Thus, outliers are but one form of misspecified model.  
We recommend a common solution for dealing with outliers and more general forms of model misspecification.}

@@

\section{Restricted Likelihood}
\label{restrictedlikelihood}
%\textcolor{blue}{Tried to make more concise. Commented out section describing the further examples related to randomized experiments, contingency tables, and meta-analysis. }

\subsection{Examples}
To describe the use of the restricted likelihood in a Bayesian framework, 
we begin with a pair of simple examples
for the one-sample problem. Several more examples are detailed in \cite{lewis2014}.  For those discussed here, the model takes the
data $\by=(y_1,\ldots,y_n)$
to be a random sample
of size $n$ from a continuous distribution indexed by a parameter
vector $\bth$, with pdf $f(y| \bth)$.  The standard, or full-data,
likelihood would be $L(\bth | \by) = \prod_{i=1}^n f(y_i | \bth)$.

As a first example, we consider the case where a subset of the data are known to be 
bad in the sense of not informing us about $\bth$--and where the subset is known.  This case mimics the setting where outliers in
a data set are identified and discarded before a formal analysis is done.  
Without loss of generality, we label the good 
cases $1$ through $n-k$ and the bad 
cases $n-k+1$ through $n$.  
The relevant likelihood to be used to move from prior distribution to posterior distribution is clearly 
$L(\bth | y_1, \ldots, y_{n-k}) = \prod_{i=1}^{n-k} f(y_i | \bth)$.  
For an equivalent analysis, we rewrite the full %entire
likelihood as the product of two pieces,
\begin{eqnarray}
\label{OutlyingCases}
L(\bth | \by)  
= \left( \prod_{i=1}^{n-k} f(y_i | \bth) \right) \left( \prod_{i=n-k+1}^{n} f(y_i | \bth) \right) .  
\end{eqnarray}
% where the second factor may not actually involve $\bth$.
We wish to keep the first piece and drop the second for better inference on $\bth$.

A second example involves deliberate censoring of small and large observations.  This is
sometimes done as a precursor to the analysis of reaction time
experiments  \citep[e.g.,][]{ratcliff1993}
\green{where very small reaction times are physiologically too quick to be true reaction times
are usually explained as anticipation, and where implausibly long reaction times are believed due to inattention. }  
%\red{Q: provide a reason why this deliberate censoring is done?}
With lower
and upper censoring times at $t_1$ and $t_2$, the post-censoring sampling distribution
is of mixed form, with masses $F(t_1|\bth)$ at $t_1$ and $1-F(t_2|\bth)$ at $t_2$,
and density $f(y | \bth)$ for $y \in (t_1, t_2)$.  We adjust the original data $y_i$,
producing $c(y_i)$ by defining $c(y_i)= t_1$ if $y_i \leq t_1$, $c(y_i)=t_2$ 
if $y_i \geq t_2$, and $c(y_i)=y_i$ otherwise.  
The adjusted update is performed with $L(\bth |c(\by))$.  
With slightly non-standard notation, we let $g(t_1|\bth) = F(t_1|\bth)$,
$g(t_2 | \bth) = 1 - F(t_2|\bth)$, and $g(y|\bth)=f(y|\bth)$ for
$y \in (t_1, t_2)$.
Hence, we may rewrite the full %complete %entire 
likelihood as the product of
two pieces, 
\begin{eqnarray}
\label{Censoring}
%L(\bth | \by) =  \left( \prod_{i=1}^n g(x_i | \bth) \right) \left( \prod_{i=1}^n f(y_i|\bth, x_i) \right) ,
L(\bth | \by) =  
\left( \prod_{i=1}^n g(c(y_i)  | \bth) \right) \left( \prod_{i=1}^n f(y_i|\bth,c(y_i) ) \right) ,  
\end{eqnarray}
and retain only the first for the formal update.

%Further examples abound. In a completely randomized experimental design, we randomize %our 
%experimental units to treatment conditions and then ignore the details of the observed randomization \citep{dean1999}; 
%in work with contingency tables,  
%we collapse categories with small counts, coarsening the scale of %our 
%data \citep{agresti2002}; in meta-analysis, we ignore the individual patient-level data and instead
%work with estimated effects from the studies \citep{orourke2007}.  
%Further examples are described in \cite{lewis2014}. 
\subsection{Generalization}

To generalize the approach in \eqref{OutlyingCases} and
\eqref{Censoring}, %, and these other settings,
we write the full likelihood in two pieces with a conditioning statistic $T(\by)$, as indicated below:
\begin{eqnarray}
\label{FullLikelihood}
L(\bth | \by)  
& = & f(T(\by) | \bth) \,\, f(\by |\bth, T(\by)) .  
\end{eqnarray}
%Throughout the paper, we use the convention that a function is defined by its inputs. 
Here,  $f(T(\by) | \bth)$ is the conditional pdf of $T(\by)$ given $\bth$ and $f(\by |\bth, T(\by))$ is the conditional pdf of $\by$ given $\bth$ and $T(\by)$.  In the dropped case example, the conditioning statistic is $T(\by) = (y_1, \ldots, y_{n-k})$.  In 
the censoring example, the conditioning statistic is $T(\by) = (c(y_1),\ldots,c(y_n))$.  We refer to 
$f(T(\by) | \bth)$ as the restricted likelihood and $L(\bth | \by)=f(\by|\bth)$ as the full likelihood.

Bayesian methods can make use of a restricted likelihood %in place of a complete likelihood
since $T(\by)$ is a well-defined random variable with a probability distribution indexed by $\bth$.  
The update from prior distribution to posterior distribution can be 
made on the basis of $f(T(\by)|\bth)$ rather than $f(\by|\bth)$.  This leads to the restricted likelihood posterior 
\begin{eqnarray}
\label{RestrictedPosterior}
\pi(\bth | T(\by)) & = & \frac{\pi(\bth) f(T(\by) | \bth)}{m(T(\by))} ,
\end{eqnarray}
where $m(T(\by))$ is the marginal distribution of $T(\by)$ under the prior distribution.  
Predictive statements for further (good) data rely on the model.  For another observation,
say $y_{n+1}$, we would have the predictive density 
\begin{eqnarray}
\label{RestrictedpredDist}
f(y_{n+1} | T(\by)) = \int f(y_{n+1} | \bth) \pi(\bth | T(\by))\ d\bth .  
\end{eqnarray}

\subsection{Literature review}
Direct use of restricted likelihood appears in many areas of the literature.  The motivation is often similar to ours:   
concern about outliers or, more generally, model misspecification.  For example, the use of rank likelihoods is discussed by \cite{savage1969}, \cite{pettitt1983, pettitt1982}, and more recently by \cite{hoff2013}.  
\green{Lewis et al. (2012 xxxx) make use of robust regression estimators for $T(\by)$.}  
Asymptotic properties of restricted posteriors are studied by \cite{doksum1990}, \cite{clarke1995}, \cite{yuan2004},  and \cite{hwang2005}. The tenor of these asymptotic 
results is that, for a variety of conditioning statistics with non-trivial regularity conditions on prior, model, and likelihood, the
posterior distribution resembles the asymptotic sampling distribution of the conditioning statistic.

%edited to include more detail on ABC:
%\textcolor{blue}{Part on the use of incomplete likelihoods being used as approximations to complete likelihoods was commented out. However, I added more detail on ABC. Thought is that we could move this to the discussion.} 
%\red{This additional discussion on ABC
%  (next two paragraphs) looks a bit too long.}
Restricted likelihoods have also been used as practical approximations to a full likelihood. For example, \cite{pratt1965} appeals to heuristic arguments regarding approximate sufficiency to justify the use of the restricted likelihood of the sample mean and standard deviation. Approximate sufficiency is also appealed to in the use of Approximate Bayesian Computation (ABC), which is related to our method.  
ABC is a collection of posterior approximation methods which has recently experienced success in applications to epidemiology, genetics, and quality control \citep[see, for example,][]{tavare1997, pritchard1999,  marjoram2003, fearnhead2012}. Interest typically lies in the full data posterior and ABC is used for computational convenience as an approximation.  Consequently, effort is made to choose an approximately sufficient $T(\by)$ and update to the ABC posterior by using the likelihood $L(\bth| \mathcal{B}(\by))$, where $\mathcal{B}(\by)=\{\by^{*}|\rho(T(\by),T(\by^{*}))<\epsilon\}$, $\rho$ is a metric, and $\epsilon$ is a tolerance level. This is the likelihood ``conditioned'' on the collection of data sets which result in a $T(\cdot)$ within $\epsilon$ of the observed $T(\by)$. %Hence, we can see that, as in our method, conditioning is done on a summary of the complete data. 
By choosing an approximately sufficient $T(\cdot)$ and a small enough $\epsilon$ the claim is made that $L(\bth|\mathcal{B}(\by))\approx L(\bth|T(\by))\approx L(\bth|\by)$. Consequently, the ABC posterior approximates the full data posterior. We note that efforts have been made to formalize what is meant by  approximate sufficiency \citep[e.g.,][]{joyce2008}.

%ABC is related to our method in that conditioning is on something other than the complete data. Even if $T(\by)$ were exactly  sufficient in the traditional sense, a positive $\epsilon$ would result in something theoretically different from the complete data posterior (even if they are indeed approximately the same). We also mention that in ABC, $\mathcal{B}(\by)$ is not a statistic in the strict sense because it does not partition the sample space before observing data. That is, a feasible data set $\bz^{*}$ can belong to more than one `ball'  $\mathcal{B}(\cdot)$.  The method relies on conditioning on this set because in most ABC applications, approximating the complete data posterior directly is hard (or impossible) and sampling data matching the observed summary exactly is also hard (or impossible). ABC methods are a practical and useful solution in these settings. For the regression models we present, sampling new data matching the observed summary exactly is indeed possible and so we can condition exactly on the observed summary. Further, we reiterate that our motivation is different from ABC in that we do not want to condition on the complete data because we assume it contains inherent flaws which can be detrimental to inference.

\green{ABC is related to our method in that the ``conditioning'' is on something other than the data $\by$.  However, the conditioning in ABC is heuristic.  For technical reasons, the calculation proceeding from $\pi(\bth)$ to $\pi(\bth)L(\bth| \mathcal{B}(\by))/\int \pi(\bth)L(\bth| \mathcal{B}(\by)) d\bth$ is not a formal conditional update.  Simply put, if an external observer told $\mathcal{B}(\by)$ can supply $T(\by)$ and vice-versa, the information contained in $\mathcal{B}(\by)$ is exactly that contained in $T(\by)$ and proper conditioning will lead to identical posterior distributions.  This is not the case with a typical implementation of ABC.}

% In a related, but distinctly different, approach which also takes advantage of summary statistics, the full data log likelihood is replaced by a loss function \citep[e.g.][]{bissiri2013} in an effort to concentrate inference only on parameters of interest. In contrast, the incomplete likelihood is formulated using a full probability model, allowing for a formal Bayesian update, while remaining robust to misspecification.

This work extends the development of Bayesian restricted likelihood by arguing \green{that deliberate choice of $T(\by)$} is sound practice and also by expanding the class of conditioning statistics in which exact conditioning can be achieved. %well beyond ranks.  
Our methods do not rely on asymptotic properties, \green{nor do they rely on} approximate conditioning, \green{nor do they need to correspond to artificial censoring to create incomplete or missing data}  \citep[e.g.,][]{albert1988, hoffwakefield2013}.  
\green{Note - fix references here hoffwakefield for asymptotic; albert for ? .}

The key to productive use of the restricted likelihood is the choice of $T(\by)$ and the development
of computational strategies that allow us to truly condition on the observed $T(\by)$ and 
fit the model in formal Bayesian fashion.  In this work, we focus
on robustness, and natural choices of $T(\by)$ for the one-sample problem include a set of middling order statistics, a trimmed
mean, or a classical robust estimator of location and/or scale.  
We have previously implemented several of these methods and have found them to perform well \citep[][]{lewis2012}. Versions most extensible to the linear model include the M-estimators in the tradition of \cite{huber1964}, least median squares (LMS), and least trimmed squares (LTS). For these choices the restricted likelihood is not available in closed form, making computation of the restricted posterior a challenge. For low-dimensional statistics $T(\by)$ and parameters $\bth$, the direct computational strategies described in \cite{lewis2014} can be used to estimate the incomplete posterior conditioned on essentially any statistic.  These strategies rely on generation of complete data sets from different values of $\bth$.  
Each complete data set leads to a statistic $T(\by)$ under $\bth$, and these generated statistics are used to estimate the density at $T(\by_{obs})$, where $\by_{obs}$ is the observed complete data. This estimate is fed into Bayes' theorem for the update from prior distribution to posterior distribution.  %For the example in the previous section, these strategies work well.
%\textcolor{blue}{Comment out `A variety of  variance reduction techniques which exploit\dots' because it seemed too vague and a bit off topic}
%In low dimensional settings (in $T(\by)$ and $\bth$) the incomplete posterior under each of these versions (and others) can be estimated by approximating the incomplete likelihood using kernel density estimation techniques \citep[for details see][]{lewis2014}.

These direct sampling techniques rely on density estimation and numerical integration. Density estimation becomes difficult for high dimensional $T(\by)$ and numerical integration breaks down for high dimensional $\bth$. For these settings, an MCMC algorithm is developed in Section \ref{highDim}. Software is available from the authors for use with simultaneous M-estimators of location and scale. At the present time, further software development is required to extend beyond simultaneous M-estimators.

%\textcolor{blue}{Added a bit about this to the discussion: Though we concentrate in this paper on applying incomplete likelihood inference for the mean of linear regression models, the computational strategies we devise in subsequent sections also allow us to apply the method to inference beyond the mean.  In particular, quantile regression falls within the framework we develop. }
%The next section develops the necessary computational strategies.

\subsection{A first data analysis example} %\textcolor{blue}{ADDED}
\label{darwinAnalysis}
Before discussing computational details of finding restricted 
posteriors, we begin by analyzing a simple example involving
outliers. The data consists of well known measurements of the \green{passage time} of
light taken by Simon Newcomb in the late nineteenth century and
contain two clear outliers on the low end. Details and data are
available in the R package {\tt MASS} and previous analyses appear in
\cite{stigler1977}, \cite{chan1980}, and \cite{gelman2004}. Letting
$y_{i}$ denote the $i^{th}$ coded measurement of the passage time of light, we
assume $y_{i}\iid N (\beta, \sigma^{2})$ for $i=1,2,\dots, n=66$. The
parameter $\beta$ can be interpreted as the passage time of light with
$\sigma^{2}$ representing measurement error. The outliers \green{are a clear indication} %make clear
that this normal model is misspecified.  \green{Also,} %Another reason for model misspecification is that 
the measurements are coarse, taking on only
integer values.  For a Bayesian analysis, we must specify prior
distributions on the parameters $\bth=(\beta, \sigma)$. For this we
assume an independent normal -- inverse gamma model. That is,
$\beta\sim N(\eta, \tau^{2})$ and $\sigma^{2}\sim IG(a, b)$. The
hyperparameters  are fixed for simplicity and taken to be $\eta=23.6$,
$\tau=2.04$, $a=5$,  and $b=10$. We note that our computational
strategy described in Section \ref{highDim} makes it feasible to fit
more complex model structures. The Bayesian model presented here is a
special case of the linear model in \eqref{LinearModel} presented below.

>>>>

% commands for for editing
\usepackage{color}
\newcommand{\red}[1]{{\color{red}#1}}
\newcommand{\blue}[1]{{\color{blue}#1}}
\newcommand{\brown}[1]{{\color{brown}#1}}
\newcommand{\green}[1]{{\color{green}#1}}
\newcommand{\magenta}[1]{{\color{magenta}#1}}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\graphicspath{{figures/}}

@@

%\newpage
\title{Bayesian Restricted Likelihood Methods}
\author{John R. Lewis, Steven N. MacEachern and  Yoonkyung Lee \\
{\small \it Department of Statistics, The Ohio State University, Columbus, Ohio 43210}\\
{\small lewis.865@osu.edu, snm@stat.osu.edu and yklee@stat.osu.edu}
\thanks{This research has been supported by Nationwide Insurance Company and by the NSF under grant numbers DMS-1007682 and DMS-1209194.  The views in this paper are not necessarily those of Nationwide Insurance or the NSF.}} 
\begin{document}
\date{}
\maketitle

\begin{abstract}
Bayesian methods have proven themselves to be successful across a wide
range of scientific problems and have many well-documented advantages
over competing methods. However, these methods run into difficulties
for two major and prevalent classes of problems: handling data sets
with outliers and dealing with model misspecification. We outline the
drawbacks of previous solutions to both of these problems (e.g., use of
heavy-tailed likelihoods) and propose the restricted likelihood as an
alternative.  When working with restricted likelihood, we summarize
the data through a set of (insufficient) statistics,
targeting inferential quantities of interest, and update the prior
distribution with the summary statistics rather than the complete
data.  By choice of conditioning statistics, we
retain the main benefits of Bayesian methods while reducing the
sensitivity of the analysis to features of the data not picked up by
the conditioning statistics. A major
contribution is the development of a data augmented MCMC algorithm
for the linear model and a wide range of choices for
summary statistics. We demonstrate the method on an
insurance agency data set containing many outliers and subject to model
misspecification. Success is manifested in better predictive
performance for data points of interest as compared to competing
methods.

@@

\section{Introduction}

Bayesian methods have provided successful solutions to a wide range of scientific problems, with their value
having been demonstrated both empirically and theoretically.  
In simple settings, the success of the methods is often attributed to formal optimality properties, sometimes derived
through the laws of subjective probability 
and sometimes through admissibility and the complete class
theorems of decision theory.  In complex settings, the hierarchical model allows one to create 
and fit sophisticated models that may, for example, pool information across similar problems.

The development of Bayesian inference relies on a complete Bayesian model consisting of 
three elements:  the prior distribution, the loss function, and the likelihood or sampling density.  While formal
optimality of Bayesian methods is unquestioned if one accepts the validity of all three of these elements, 
 a healthy skepticism encourages us to question each of them.  Concern about the prior distribution has
been addressed through the development of techniques for subjective elicitation \citep{garthwaite2005} and
the rise of objective Bayesian methods \citep{berger2006}.  
Concern about the loss function is reflected in, for example, the
extensive literature on Bayesian hypothesis tests \citep{kass1995}.  
The sampling density has been given less attention from a specifically
Bayesian view, although the work on predictive diagnostics \citep{box1980} departs from classical
traditions.

The focus of this work is the intersection of Bayesian methodology and data analysis.  In particular, 
we develop techniques to handle imperfections in the sampling density.  These imperfections often show
themselves through the presence of outliers--here taken to be cases not reflecting the phenomenon under
study--in the data set.  
There are three main solutions to Bayesian outlier-handling.  The first is to replace the basic sampling
density with a mixture model which includes one component for the ``good'' data and a second 
component for the ``bad'' data.  With this approach, the prior distribution is updated with the likelihood
from the mixture model to obtain the complete-data posterior distribution.  The good component of
the sampling density is used for prediction of future good data.  The second approach replaces the
basic sampling density with a thick-tailed density in an attempt to discount outliers, yielding techniques that 
often provide solid estimates of the center of the distribution but 
do not easily translate to predictive
densities for further good data.  The third approach fits a flexible (typically nonparametric) model to 
the data, producing a Bayesian version of a density estimate for both good and bad data.  In recent 
development, inference 
is made through the use of robust inference functions \citep[][in press]{lee2013}.

The traditional strategies for handling outliers all have their drawbacks.  While we view the sampling 
density for the good data as stable, the outlier-generating processes 
may be transitory in nature, constantly shifting as the source of bad data changes.  This prevents
us from appealing to large-sample arguments to claim that, with enough data, we can nail down a
model for both good and bad data combined.  Instead of attempting to model both good and bad
data, we propose a novel strategy for handling outliers:  the use of restricted likelihood.  In a nutshell,
we begin with a complete model   
as if all of the data are good.
But rather than driving the move from prior distribution to posterior
distribution by the entire likelihood, we use only the likelihood of a
few summary statistics which typically target inferential quantities
of interest.  We call this reduced likelihood a restricted likelihood.  The 
update is a formal update from prior distribution to posterior distribution, based on the sampling density
of the summary statistics. 
 In our approach, the reader will identify a stream of reasoning which is manifested in
classical M-estimation, generalized estimating equations, approximate Bayesian computation,
and elsewhere.  The novelty of the work is twofold.  We make use of
classical robust estimators as  summary statistics in a formal
Bayesian framework, using the sampling density of the estimators as a replacement for the 
sampling density of the data.  We advance the argument that conditioning on an insufficient
summary of the data is sound practice, rather than merely being done for computational and 
modelling convenience.

In addition to outlier-prone data, the sampling density can err due to model misspecification.  
The traditional view is that, if the model is inadequate, one should build a better model.  In our
empirical work, as data sets have become larger and more complex, we have bumped into 
settings where we cannot realistically build the perfect model.  We ask the question ``by attempting
to improve our model through elaboration, will the overall performance of the model suffer?''  If yes,
we avoid the elaboration, retaining a model with some level of misspecification.  Acknowledging that
the model is misspecified implies acknowledging that the sampling density is incorrect, exactly as
we do when outliers are present.  In this sense, misspecified models and outliers are reflections of the
same phenomenon, and we recommend a common solution for dealing with the problem.

@@

\section{Restricted Likelihood}
\label{restrictedlikelihood}

To describe the use of restricted likelihood in a Bayesian framework, 
we begin with a pair of simple examples
for the one-sample problem.  In each, the model takes the
data $\by=(y_1,\ldots,y_n)$
to be a random sample
of size $n$ from a continuous distribution indexed by a parameter vector $\bth$.  The standard, or complete,
likelihood would be $L(\bth | \by) = \prod_{i=1}^n f(y_i | \bth)$.

As a first example, we consider the case where a subset of the data are known to be 
bad in the sense of not
  informing us about $\bth$--and where the subset is known.  This case mimics the setting where outliers in
a data set are identified and discarded before a formal analysis is done.  
Without loss of generality, we label the good 
cases $1$ through $n-k$ and the bad 
cases $n-k+1$ through $n$.  
The relevant likelihood to be used to move from prior distribution to posterior distribution is clearly 
$L(\bth | y_1, \ldots, y_{n-k}) = \prod_{i=1}^{n-k} f(y_i | \bth)$.  
For an equivalent analysis, we rewrite the entire likelihood as the product of two pieces,
\begin{eqnarray}
\label{OutlyingCases}
L(\bth | \by)  
= \left( \prod_{i=1}^{n-k} f(y_i | \bth) \right) \left( \prod_{i=n-k+1}^{n} f(y_i | \bth) \right) ,
\end{eqnarray}
where the second term may not actually involve $\bth$.
We wish to keep the first piece and drop the second for better inference on $\bth$.

A second example involves deliberate censoring of small and large observations.  This is
sometimes done as a precursor to the analysis of reaction time experiments (e.g., \cite{ratcliff1993}).  
With lower
and upper censoring times at $t_1$ and $t_2$, the post-censoring sampling distribution
is of mixed form, with masses $F(t_1|\bth)$ at $t_1$ and $1-F(t_2|\bth)$ at $t_2$,
and density $f(y | \bth)$ for $y \in (t_1, t_2)$.  We adjust the original data $y_i$,
producing $x_i$ by defining $x_i = t_1$ if $y_i \leq t_1$, 
$x_i = t_2$ if $y_i \geq t_2$, and $x_{i}=y_{i}$ otherwise.  
The adjusted update is performed with $L(\bth | \bx)$. 
With slightly non-standard notation, we let $g(t_1|\bth) = F(t_1|\bth)$,
$g(t_2 | \bth) = 1 - F(t_2|\bth)$, and $g(y|\bth)=f(y|\bth)$ for
$y \in (t_1, t_2)$.
Alternatively, we may rewrite the entire likelihood as the product of
two pieces, 
\begin{eqnarray}
\label{Censoring}
L(\bth | \by) =  
\left( \prod_{i=1}^n g(x_i | \bth) \right) \left( \prod_{i=1}^n f(y_i|x_i,\bth) \right) ,  
\end{eqnarray}
and retain only the first for the formal update.

Further examples abound.  In a completely randomized experimental design, we randomize %our 
experimental units to treatment conditions and then ignore the details of the observed randomization \citep{dean1999}; 
in work with contingency tables,  
we collapse categories with small counts, coarsening the scale of %our 
data \citep{agresti2002}; in meta-analysis, we ignore the individual patient-level data and instead
work with estimated effects from the studies \citep{orourke2007}.  
Further examples are described in \cite{lewis2014}.  
To describe the approach in (\ref{OutlyingCases}), 
(\ref{Censoring}), and these other settings,
we write the complete data likelihood in two pieces, as indicated below:
\begin{eqnarray}
\label{FullLikelihood}
L(\bth | \by)  
& = & f(T(\by) | \bth) \,\, f(\by | T(\by), \bth) .  
\end{eqnarray}
In the dropped case example, the conditioning statistic is $T(\by) = (y_1, \ldots, y_{n-k})$.  In 
the censoring example, the conditioning statistic is $T(\by) = (x_1, \ldots, x_n)$.  We refer to 
$f(T(\by) | \bth)$ as the restricted likelihood.

Bayesian methods can make use of restricted likelihood in place of the complete data likelihood
since $T(\by)$ is a well-defined random variable with a probability distribution indexed by $\bth$.  
The update from prior distribution to posterior distribution is 
made on the basis of $f(T(\by)|\bth)$ rather than $f(\by|\bth)$.  This leads to the restricted likelihood posterior 
\begin{eqnarray}
\label{RestrictedPosterior}
\pi(\bth | T(\by)) & = & \frac{\pi(\bth) f(T(\by) | \bth)}{m(T(\by))} ,
\end{eqnarray}
where $m(T(\by))$ is the marginal distribution of $T(\by)$ under the prior distribution.  
Predictive statements for further (good) data rely on the model.  For another observation,
say $y_{n+1}$, we would have the predictive density 
\[
f(y_{n+1} | T(\by)) = \int f(y_{n+1} | \bth) \pi(\bth | T(\by))\ d\bth .  
\]

Direct use of restricted likelihood appears in many areas of the literature.  The motivation is often similar to ours:   
concern about model misspecification.  For example, the use of rank likelihoods is discussed by \cite{savage1969}, \cite{pettitt1983, pettitt1982}, and more recently by \cite{hoff2013}. Asymptotic properties of restricted posteriors are studied by \cite{doksum1990}, \cite{clarke1995}, \cite{yuan2004},  and \cite{hwang2005}. The tenor of these asymptotic 
results is that, for a variety of conditioning statistics with non-trivial regularity conditions on prior, model, and likelihood, the
posterior distribution resembles the asymptotic sampling distribution of the conditioning statistic.

Restricted likelihoods have also been used as practical approximations to the full likelihood. For example, \cite{pratt1965} appeals to heuristic arguments regarding approximate sufficiency to justify the use of the restricted likelihood of the sample mean and standard deviation. Approximate sufficiency is  often appealed to when applying approximate Bayesian computation (ABC), a collection of posterior approximation methods which has recently experienced success in applications to epidemiology, genetics, and quality control \citep[see, for example,][]{tavare1997, pritchard1999,  marjoram2003, fearnhead2012}. ABC is typically used when likelihoods are intractable but simulation of new data from the model is easily done. In the cited references, interest lies in the full data posterior and ABC is used for computational 
convenience.  Consequently, effort is made to choose a $T(\by)$ such that the likelihood 
$L(\bth|\by) \approx L(\bth|T(\by))$.  Technical limitations of ABC imply that in realistic (and all non-discrete) settings,
the conditioning is not exact, but approximate.

In a related, but distinctly different, approach which also takes advantage of summary statistics, the full data log likelihood is replaced by a loss function \citep[e.g.][]{bissiri2013} in an effort to concentrate inference only on parameters of interest. In contrast, the restricted likelihood is formulated using a full probability model, allowing for a formal Bayesian update, while remaining robust to misspecification. This work extends the use of restricted likelihood by arguing that its use is sound practice, and also by expanding the class of conditioning statistics in which exact conditioning can be achieved well beyond ranks.  Our methods do not rely on asymptotic properties or approximate conditioning as in previous work (e.g., \cite{albert1988},\cite{hoffwakefield2013}).

The key to productive use of restricted likelihood is the choice of $T(\by)$ and the development
of computational strategies that allow us to truly condition on the observed $T(\by)$ and 
fit the model in formal Bayesian fashion.  In this work, we focus
on robustness, and natural choices of $T(\by)$ include a set of middling order statistics, a trimmed
mean, or a classical robust estimator of location and/or scale.  
We have previously implemented all of these methods for one-sample problems 
where we have found them to perform well (e.g., \cite{lewis2012}).  Of these versions, the most 
extensible to the linear model are the
M-estimators, 
in the tradition of \cite{huber1964}.  
The computational strategies we devise in subsequent sections allow us to apply Bayesian
restricted likelihood inference well beyond regression models for the mean of a distribution.  In particular, 
quantile regression falls within the framework we develop.
The next section develops the necessary computational
strategies.
